Demes: a standard format for demographic models

### 2022-04-21 paper.tex
add barton et al 2010
misc edits

\title{Demes: a specification for interchanging complex demography}
% First authors
\author[1,$\star$]{Graham Gower}
\author[2,$\star$]{Aaron P. Ragsdale}

@@

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being estimated.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
graphical model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.
\end{abstract}

@@

Simulation is a core tool in population genetics, and a
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.,][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.,][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.,][]{
thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,
becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
However, there is no standard format shared across simulators,
and the details of the demographic models are not separated
from other simulation parameters.
The APIs used by individual simulators to define demographic models are also
not appropriate as an interchange standard because of the dependency on both
the programming language involved
and the specific details of the simulator in question.

@@

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The demographic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\citep{wright1943isolation,barton2002neutral,ringbauer2017inferring,battey2020space}
or stochastic population size dynamics
\citep{nunney1993assessing,orr2014population},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
The basic assumptions of discrete populations connected by instantaneous or continuous
migrations are also shared by many inference methods
\citep[e.g.,][]{gutenkunst2009inferring,gravel2012population,kamm2017efficient,
jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
Demes therefore serves as ``middleware'' between inference methods and simulation
software, capturing common assumptions shared across implemented software.
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

@@

While the high-level specification is designed for human readability and
conciseness, the underlying data model suitable for software implementation
is redundant and exhaustive. 
Translation from the high-level specification to the low-level data model
requires resolving heierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as an input.
By having relatively few high-quality Demes parsers available as libraries
(ideally one per programming language), we ensure consistency across simulation
software in their implementation of Demes-specified models.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs.
In addition, we have high-quality parser implementations in the Python and
C languages (published under liberal open source licences), providing a
solid foundation for the software ecosystem (Table~\ref{tab:software}).

@@

The Demes specification is static by design -- we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{fig/showcase}
    \caption{
        \textbf{Illustration and simulation using Demes.}
        (A)
        Using an inferred demographic model from \citet{tennessen2012evolution}
        specified as a YAML in Demes format (Figure~\ref{fig:tennessen}), we
        used \demesdraw\ to visualize the demographic model (note the recent
            exponential growth resulting in present-day population sizes that
        greatly exceed those in the past).
        We then used \msprime\ to simulate genomic data for 20 genome copies
        sampled from the two contemporary populations, and we used \moments\
        to compute the expected joint site-frequency spectrum for the same
        sample sizes (Figure~\ref{fig:tennessen-simulation}).
        (B, C) We compared the single-population SFS in each population, showing
        agreement between the simulation methods.
        (D) Code snippets of the interactions between \demes\ and the simulation
        software. An extended script to compute the SFS shown in (B) and (C) is
        given in Figure~\ref{fig:tennessen-simulation}.
    }
    \label{fig:showcase}
\end{figure}

@@

The difficulty of describing complex demographic models
for population genetic simulations has long been acknowledged
~\citep[e.g.,][]{antao2007modeler4simcoal2}, % jk - any older refs for this?
and a number of different methods have been proposed to
make such simulations more accessible and less error prone.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
ewing2010msms,zhou2018popdemog}. However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the specific syntax
of a given simulator. Another approach that has been taken to
simplify running such simulations is to provide an API in a
scripting language such as Python or R, providing a flexible
interface either to a built-in simulation
engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,
kelleher2016efficient,haller2017flexible}
% jk: are/were there other simulation frontends out there?
or as a front end for other simulators~\citep{staab2016coala}.
While APIs have many advantages and have grown in popularity in
recent years, the demographic model descriptions are tied
to the details of both the underlying simulation engines
and the programming language in question.

@@

Population sizes are given as numbers of individuals, and details
such as ploidy levels are considered external to the model.
We therefore focus on the number of individuals as opposed
to the number of genome copies.
Sizes and mating system details are specified for each population within
epochs.
Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

@@

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure), as are
sequence length, annotations and other genomic features, and individual ploidy.
Selection and dominance models are absent, as discussed in
Sec.~\ref{sec:appendix-pop-gen-model}.
We also omit information about the sampling of individuals, as we view
the demographic history as independent of the sampling of individuals.
It is therefore left up to simulation and inference tools how to define
genomic parameters and sampling strategies, as each software will have its
own set of constraints and relevant input parameters.

@@

\begin{figure}[h!]
    \begin{tcolorbox}
        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yml}
    \end{tcolorbox}
    \caption{
        \textbf{The \citet{tennessen2012evolution} two-population demographic model in Demes format.}
    }
    \label{fig:tennessen}
\end{figure}

\begin{figure}[h!]
    \begin{tcolorbox}
        % NOTE: line highlights may need adjusting after changing this code
        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt,highlightlines={9,19,52}]{python}{models/tennessen-simulation.py}
    \end{tcolorbox}
    \caption{
        \textbf{Simulation of SFS for the Tennessen model.}
        We first load the demographic model using \demes\ (as \texttt{graph}),
        which can then be used by \msprime\ to create the demographic model used in
        \texttt{msprime.sim\_ancestry()}. The same loaded graph can also be
        passed to \moments\ to computed the expected joint SFS.
        To compare the SFS in Figure~\ref{fig:showcase}, we \emph{marginalize} the
        joint SFS to obtain the single-population SFS for both AFR and EUR populations.
        Lines interfacing \demes\ and other software are highlighted.
    }
    \label{fig:tennessen-simulation}
\end{figure}

>>>>

\title{Demes: a standard format for demographic models}
% First authors
\author[1,$\star$]{Graham Gower}
\author[2,$\star$]{Aaron P. Ragsdale}

@@

\begin{abstract}
Understanding the demographic history of populations is a
key goal in population genetics, and with improving methods
and data, ever more complex models are being constructed.
Demographic models of current interest
typically consist of a set of discrete populations,
their sizes and growth rates, and continuous and pulse migrations
between those populations over a number of epochs, which can require
dozens of parameters to fully describe. There is currently
no standard format to define such models, significantly
hampering  progress in the field. In particular, the important
task of translating the model descriptions in published
work into input suitable for population genetic simulators is labour intensive
and error prone.
We propose the Demes data model and file format,
built on widely used technologies,
to alleviate these issues. Demes provides a well-defined and unambiguous
model of populations and their properties that is straightforward to
implement in software, and a text file format that is designed for
simplicity and clarity.
We provide thoroughly tested implementations of Demes parsers in Python
and C, and showcase initial support in a number of simulators and inference
methods.

\ggcomment{We don't showcase Demes output from any inference software.}
\end{abstract}

@@

Simulation is a core tool in population genetics, and a
many methods have been developed over the past three
decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models,
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approaches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.,][]{hudson2002generating,hernandez_flexible_2008,kern2016discoal},
a custom input file
format~\citep[e.g.,][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models to be defined programmatically~\citep[e.g.,][]{
thornton2014cpp,thornton2019-nu,baumdicker2021-iu,kelleher2016efficient,
becheler2019quetzal,haller2019slim,hernandez_sfs_code_2015}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.,][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation software~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and allow for documentation in the form of comments.
However, there is no standard format shared across simulators,
and the details of the demographic models are not separated
from other simulation parameters.
The APIs used by individual simulators to define demographic models are also
not appropriate as an interchange standard because of the dependency on both
the programming language involved
and the specific details of the simulator in question.

@@

Demographic models consist of one or more populations (or ``demes'') defined
by their size histories and the time intervals of their existence.
Individuals can move between populations based on their ancestor-descendant
relationships (i.e., the topology) or by continuous or discrete migration events.
The demographic model itself omits information about genome biology, such as
genome size and architecture, ploidy, and mutation and recombination rates.
It also omits any information about selection operating within populations.
Within populations, allele-frequency dynamics are assumed to
follow the Wright-Fisher model, so that generations are discrete and
non-overlapping and population sizes are independent of mean fitness
(i.e., ``soft'' selection~\citep{christiansen1975hard}).
While these restrictions preclude some population-genetic scenarios of
interest, such as continuous spatial structure
\citep{wright1943isolation,barton2002neutral,barton2010new,
ringbauer2017inferring,battey2020space}
or stochastic population size dynamics
\citep{nunney1993assessing,orr2014population},
many current forwards- and backwards-in-time simulation software
make these same assumptions, so that Demes is relevant to a range of
widely-used simulators.
The basic assumptions of discrete populations connected by instantaneous or continuous
migrations are also shared by many inference methods
\citep[e.g.,][]{gutenkunst2009inferring,gravel2012population,kamm2017efficient,
jouganous2017inferring,ragsdale2019models,excoffier2021fastsimcoal2}.
Demes therefore serves as ``middleware'' between inference methods and simulation
software, capturing common assumptions shared across implemented software.
In Appendix~\ref{sec:appendix-pop-gen-model}, we expand on these concepts
and provide more details on the population genetics models supported by Demes.

@@

\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-parsers}}

While the high-level specification is designed for human readability and
conciseness, the underlying data model suitable for software implementation
is redundant and exhaustive. 
Translation from the high-level specification to the low-level data model
requires resolving hierarchically-defined default values and verifying
relationships between populations and the validity of specified parameter
values.
Because this translation and validation requires significant
programming effort, we define a standard software entity as part of the
specification to perform this task (the parser),
which is intended to be shared by programs that support Demes as an input.
By having relatively few high-quality Demes parsers available as libraries
(ideally one per programming language), we ensure consistency across simulation
software in their implementation of Demes-specified models.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs.
In addition, we have high-quality parser implementations in the Python and
C languages (published under liberal open source licences), providing a
solid foundation for the software ecosystem (Table~\ref{tab:software}).

@@

\ggcomment{TODO: deduplicate this section and Appendix~\ref{sec:appendix-scope}}

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{fig/showcase}
    \caption{
        \textbf{Illustration and simulation using Demes.}
        (A)
        Using an inferred demographic model from \citet{tennessen2012evolution}
        specified as a YAML file in Demes format (Figure~\ref{fig:tennessen}), we
        used \demesdraw\ to visualize the demographic model (note the recent
            exponential growth resulting in present-day population sizes that
        greatly exceed those in the past).
        We then used \msprime\ to simulate genomic data for 20 genome copies
        sampled from the two contemporary populations, and we used \moments\
        to compute the expected joint site-frequency spectrum for the same
        sample sizes (Figure~\ref{fig:tennessen-simulation}).
        (B, C) We compared the single-population SFS in each population, showing
        agreement between the simulation methods.
        (D) Python code snippets of the interactions between \demes\ and the simulation
        software. An extended script to compute the SFS shown in (B) and (C) is
        given in Figure~\ref{fig:tennessen-simulation}.
    }
    \label{fig:showcase}
\end{figure}

@@

The difficulty of describing complex demographic models
for population genetic simulations has long been acknowledged
~\citep[e.g.,][]{antao2007modeler4simcoal2}, % jk - any older refs for this?
and a number of different methods have been proposed to
make such simulations more accessible and less error prone.
Several Graphical User Interface (GUI) and visualisation
methods have been developed, which greatly facilitate
interpretation~\citep{mailund2005coasim,antao2007modeler4simcoal2,
parreira_spams_2009,ewing2010msms,parobek_skelesim_2017,zhou2018popdemog}.
However, these methods
currently have little traction as they are all either directly coupled
to an internal simulation method or to the specific syntax
of a given simulator. Another approach that has been taken to
simplify running such simulations is to provide an API in a
scripting language such as Python or R, providing a flexible
interface either to a built-in simulation
engine~\citep{thornton2014cpp,thornton2019-nu,baumdicker2021-iu,
hernandez_sfs_code_2015,kelleher2016efficient,haller2017flexible}
% jk: are/were there other simulation frontends out there?
or as a front end for other simulators~\citep{staab2016coala}.
While APIs have many advantages and have grown in popularity in
recent years, the demographic model descriptions are tied
to the details of both the underlying simulation engines
and the programming language in question.

@@

Population sizes are given as numbers of individuals, and details
such as ploidy levels are considered external to the model.
We therefore focus on the number of individuals as opposed
to the number of genome copies.
Sizes and mating system details are specified for each population within
population-specific epochs.
Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

@@

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure), as are
sequence length, annotations and other genomic features, and individual ploidy.
Selection and dominance models are absent, as discussed in
Sec.~\ref{sec:appendix-pop-gen-model}.
We also omit information about the sampling of individuals, as we view
the demographic history as independent of the sampling of individuals.
It is therefore left up to simulation and inference tools how to define
genomic parameters and sampling strategies, as each software will have its
own set of constraints and relevant input parameters.

@@

\begin{figure}[h!]
    \begin{tcolorbox}
        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/tennessen.yaml}
    \end{tcolorbox}
    \caption{
        \textbf{The \citet{tennessen2012evolution} two-population demographic model in Demes format.}
    }
    \label{fig:tennessen}
\end{figure}

\begin{figure}[h!]
    \begin{tcolorbox}
        % NOTE: line highlights may need adjusting after changing this code
        \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt,highlightlines={9,19,52}]{python}{models/tennessen-simulation.py}
    \end{tcolorbox}
    \caption{
        \textbf{Simulation of SFS for the Tennessen model.}
        We first load the demographic model using \demes\ (as \texttt{graph}),
        which can then be used by \msprime\ to create the demographic model used in
        \texttt{msprime.sim\_ancestry()}. The same loaded graph can also be
        passed to \moments\ to compute the expected joint SFS.
        To compare the SFS in Figure~\ref{fig:showcase}, we \emph{marginalize} the
        joint SFS to obtain the single-population SFS for both AFR and EUR populations.
        Lines interfacing \demes\ and other software are highlighted.
    }
    \label{fig:tennessen-simulation}
\end{figure}

### 2022-04-21 software-table.tex
misc edits

\toprule
\multicolumn{2}{l}{Software infrastructure}\\
\midrule
\texttt{demes-python} &
    A Python library for loading, saving, and working with
    Demes models. Includes conversion utilities for existing
    demographic model descriptions such as
    \texttt{ms}~\citep{hudson2002generating}
    (\url{https://github.com/popsim-consortium/demes-python}).\\

@@

\dadi &
    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
    \\

\fwdpy &
    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}\\

@@

\stdpopsim &
    A community-maintained collection of empirical data for species
    and demographic models specified in Demes
    format~\citep{adrion2020community}.\\

>>>>

\toprule
\multicolumn{2}{l}{Software infrastructure}\\
\midrule
\texttt{demes-python} &
    A Python library for loading, saving, and working with
    Demes models. Includes support for converting to and from
    \texttt{ms}~\citep{hudson2002generating}
    (\url{https://github.com/popsim-consortium/demes-python}).\\

@@

\dadi &
    Summary of what dadi does and how it implements Demes~\citep{gutenkunst2009inferring}.
    \ggcomment{what is the status of dadi/Demes support?}
    \\

\fwdpy &
    Forward-time Wright-Fisher simulation \citep{thornton2014cpp,thornton2019-nu}.\\

@@

\stdpopsim &
    A community-maintained collection of empirical data for species
    and demographic models specified in Demes
    format~\citep{adrion2020community}.
    \ggcomment{stdpopsim doesn't yet have Demes support}
    \\
